---
title: "Data exploration"
author: Marina Kochuten
date: 2-4-25
format:
  html:
    embed-resources: true
messages: false
warning: false
editor_options: 
  chunk_output_type: console
 
---

## About

In this workflow, I am exploring the [plant pollinator interaction data at HJ Andrews Experimental Forest from 2011 to 2021](https://portal.edirepository.org/nis/metadataviewer?packageid=knb-lter-and.5216.8). At this point, I am working to familiarize myself with the data, do any necessary cleaning and wrangling, and explore potential trends through a handful of preliminary visualizations.

## Setup

```{r}
# Load libraries
library(tidyverse)
library(janitor)
library(here)

# Read in pollinator data
pp_interactions <- read_csv(here("data", "pp-interactions.csv")) |>
    clean_names()

# Not sure yet if I will use this data set, but this is where flower abundance data is kept
flower_surveys <- read_csv(here("data", "flower-surveys.csv")) |>
    clean_names()
```

## Exploration

### Structure of data
```{r}
# How many observations indicate pollination interactions?
print(paste("Number of pollinator interactions incidents in df:", count(pp_interactions, ppi_status)))


print(count(pp_interactions, year))
# Missing data in 2019 and 2020, limited counts in 2021.
# I think I should filter to 2011-2018
```


### Pollination over time
```{r}
# Let's take a look at the average number of pollinator visits each year
pp_interactions|>
    filter(year %in% c(2011:2018)) |>
    group_by(year) |>
    summarise(no_int = mean(no_int, na.rm = TRUE)) |>
    ungroup() |> 
ggplot(aes(x = year, y = no_int)) +
    geom_point() +
    geom_line() +
    labs(title = "Average number of pollinator visits each year across all sites",
         subtitle = "Sharp decline from 2013 to 2015, steadily increasing")
```

```{r}
# Let's also look at the total number of pollinator visits
pp_interactions|>
    filter(year %in% c(2011:2018)) |>
    group_by(year) |>
    summarise(no_int = sum(no_int, na.rm = TRUE)) |>
    ungroup() |> 
ggplot(aes(x = year, y = no_int)) +
    geom_point() +
    geom_line() +
    labs(title = "Total number of pollinator visits each year across all sites",
         subtitle = "Increasing since 2015, due sampling effort, or a true trend?")
```

```{r}
# Look at hourly pollination trends
pp_interactions |>
    filter(year %in% c(2011:2018)) |>
    group_by(start_time) |>
    summarise(no_int = mean(no_int, na.rm = TRUE)) |>
    ungroup() |>
    ggplot(aes(x = start_time, y = no_int)) +
    geom_point() +
    labs(title = "Average number of pollinator visits by time of day")
```


### Pollination and weather

I'm curious to see how number of pollinator interactions vary under different weather conditions. My data contains the columns wind, clouds, and temp. Let's take a look at those:
```{r}
#| fig-cap: "There are many more pollinator interactions observed when it is still (ST) rather than breezy (BR)"

# Wind vs # of pollinator interactions
pp_interactions |>
    group_by(wind) |>
    summarise(interactions = sum(no_int, na.rm=TRUE)) |>
    ungroup() |>
    ggplot(aes(x = wind, y = interactions)) +
    geom_col() +
    labs(title = "Wind vs # pollinator interactions")
```

```{r}
#| fig-cap: "As expected, there are many more pollinator interactions observed when it is sunny, and interactions decrease as clouds increase"
# Clouds vs # of pollinator interactions
pp_interactions |>
    group_by(clouds) |>
    summarise(interactions = sum(no_int, na.rm=TRUE)) |>
    ungroup() |>
    mutate(clouds = fct_reorder(clouds, interactions)) |>
    ggplot(aes(x = clouds, y = interactions)) +
    geom_col() +
    coord_flip()
```

```{r}
#| fig-cap: "Seems like there is a sweet spot for pollinators around 25 degrees. I wonder if this is because of average temp when the most flowers are blooming?"
# Temp vs # of pollinator interactions
pp_interactions |>
    group_by(temp) |>
    summarise(interactions = sum(no_int, na.rm=TRUE)) |>
    ungroup() |>
    ggplot(aes(x = temp, y = interactions)) +
    geom_point()
```

### Pollinator type

Now, I'd like to explore potential trends regarding pollinator type. First, I want to group the specific pollinator species into generic types:
```{r}
# Add a column with pollinator generic type
pp_interactions <- pp_interactions |>
    mutate(pol_type = case_when(str_detect(vissp_type, "beetle") ~ "Beetle",
                                str_detect(vissp_type, "fly") ~ "Fly",
                                str_detect(vissp_type, "wasp") ~ "Wasp",
                                str_detect(vissp_type, "butterfly") ~ "Butterfly",
                                str_detect(vissp_type, "grasshopper") ~ "Grasshopper",
                                str_detect(vissp_type, "moth") ~ "Moth",
                                str_detect(vissp_type, "hummingbird") ~ "Hummingbird",
                                str_detect(vissp_type, "bee") ~ "Bee",
                                str_detect(vissp_type, "ant") ~ "Ant",
                                str_detect(vissp_type, "spider") ~ "Spider",
                                .default = "other"))

```

Now I can poke around with potential pollinator type trends

```{r}
#| fig-cap: "There is not any clearly visible trend regarding number of bee interactions over the 2018 season"
# Number of bee interactions over the 2018 field season
pp_interactions |>
    filter(year == 2018) |>
    filter(pol_type == "Bee") |>
    group_by(pol_type, sampledate) |>
    summarise(no_int = mean(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = sampledate, y = no_int)) +
    labs(title = "Number of bee interactions over the 2018 field season") +
    geom_point()
```

```{r}
#| fig-cap: "Number of bees observed is cyclical, with a huge year in 2018. Important to note that in 2021 only visitations by bees were recorded rather than visitations by all pollinators, so this year was removed."
# Average number of bee interactions over time?
pp_interactions |>
    filter(year != 2021) |>
    filter(pol_type == "Bee") |>
    group_by(year) |>
    summarise(total_bee = sum(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = year, y = total_bee)) +
    geom_line() +
    labs(title = "Number of bee interactions") 
```

```{r}
# Average number of wasp interactions over time?
pp_interactions |>
    filter(year != 2021) |>
    filter(pol_type == "Wasp") |>
    group_by(year) |>
    summarise(total_wasp = sum(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = year, y = total_wasp)) +
    geom_line() +
    labs(title = "Number of Wasp interactions") 
```

```{r}
# Average number of ant interactions over time?
pp_interactions |>
    filter(year != 2021) |>
    filter(pol_type == "Ant") |>
    group_by(year) |>
    summarise(total_ant = sum(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = year, y = total_ant)) +
    geom_line() +
    labs(title = "Number of Ant interactions") 
```

```{r}
# Average number of fly interactions over time?
pp_interactions |>
    filter(year != 2021) |>
    filter(pol_type == "Fly") |>
    group_by(year) |>
    summarise(total_fly = sum(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = year, y = total_fly)) +
    geom_line() +
    labs(title = "Number of Fly interactions") 
```

```{r}
# Average number of beetle interactions over time?
pp_interactions |>
    filter(year != 2021) |>
    filter(pol_type == "Beetle") |>
    group_by(year) |>
    summarise(total_beetle = sum(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = year, y = total_beetle)) +
    geom_line() +
    labs(title = "Number of Beetle interactions") 
```

```{r}
# Average number of grasshopper interactions over time?
pp_interactions |>
    filter(year != 2021) |>
    filter(pol_type == "Grasshopper") |>
    group_by(year) |>
    summarise(total_gh = sum(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = year, y = total_gh)) +
    geom_line() +
    labs(title = "Number of Grasshopper interactions") 
```

```{r}
# Average number of moth interactions over time?
pp_interactions |>
    filter(year != 2021) |>
    filter(pol_type == "Moth") |>
    group_by(year) |>
    summarise(total_moth = sum(no_int, na.rm = TRUE)) |>
    ggplot(aes(x = year, y = total_moth)) +
    geom_line() +
    labs(title = "Number of Moth interactions") 
```

Haha, that was fun. 

### Pollinator types and specific flower species

It could be interesting to do some research to identify important or noteworthy meadow flower species and see which polliators they are most visited by.
```{r}
# could see pollinator types by plant species - potentially identify important species?
pp_interactions |>
    group_by(pltsp_code, pol_type) |>
    summarise(no_int = sum(no_int, na.rm = TRUE)) |>
    filter(pltsp_code == "ERIGFOLI") |>
    ggplot(aes(x = pol_type, y = no_int)) +
    geom_col() +
    labs(title = "Number of interactions by pollinator for plant code ERIGFOLI")
```

### Important plants for specific pollinators

```{r}
# Explore most important plant species for bees?
pp_interactions |>
    group_by(pltsp_code, pol_type) |>
    summarise(no_int = sum(no_int, na.rm = TRUE)) |>
    ungroup() |>
    filter(pol_type == "Bee") |>
    mutate(pltsp_code = fct_reorder(pltsp_code, no_int)) |>
    arrange(desc(no_int)) |>
    slice_head(n = 10) |>
    ggplot(aes(x = pltsp_code, y = no_int)) +
    geom_col() +
    labs(title = "Top 10 plants most visited by bees") +
    coord_flip()
# The biggest one is Gilia capitata: "Globe gilia is very attractive to honey bees"
```

- Does the number of plant-pollinator interactions change over time? Does it vary by pollinator type (flies, bees, beetles, etc.)?
    - interactions across seasons and years, as well as hourly interactions.
    
- What time of day are each pollinator type more often observed?

- Does the number of plant-pollinator interactions vary with weather conditions (clouds, wind, temp)?
- Are certain pollinator types (flies, bees, beetles, etc.) more likely to interact with plants under certain weather conditions?

## Answering Part 2c Questions

1. **What have you learned about your data? Have any potentially interesting patterns emerged?**

- Originally, I thought that I had 10 years of data (such a nice round number), however surveys did not happen in 2019 nor 2020 (understandably) and 2021 surveys were for bees only, leaving me with 7 years of fairly comparable data. 
- I am not sure that I have found any potentially interesting patterns so far, however I have come up with a few more questions and perhaps more direction for my final viz.

2. **In HW #1, you outlined some questions that you wanted to answer using these data. Have you made any strides towards answering those questions? If yes, how so? If no, what next steps do you need to take (e.g. I need to create X plot type, I still need to track down Y data, I need to restructure existing data so that you can visualize it in Z ways, etc.)?**

-  Exploring the question, does the number of plant-pollinator interactions vary with weather conditions (clouds, wind, temp), resulted with a pretty intuitive outcome. More interactions were recorded when it was sunny and still.
- I looked into very general trends regarding the number of plant-pollinator interactions over the entirety of my data set, interactions by time of day, and briefly into trends during season (just for bees in 2018). There seems to be a cyclical nature of pollination visits over the entirety of the data set, and I am curious to explore potential underlying meaning, particularly plotting number of interactions against flower abundance. This will require me to dig into the flower survey data, where the authors have indicated that number of flower stalks are a good estimator of abundance. Just from a quick look, I do not see any obvious trends in pollination by time of day. 
- I ended up exploring a bit of what plants see the highest visits from specific pollinator types, and ended up with the top 10 plants visited by bees. I think this could be an interesting place to take my project, especially if I narrowed my focus to a few key pollinators, or perhaps even a focus on key flower species and exploring what pollinators are most important to them. 

3. **What challenges do you foresee encountering with your data? These can be data wrangling and / or visualization challenges.**

- My data is intuitive and quite complete, which is great! I do not expect to have many issues with wrangling or other underlying issues. I think the biggest challenge I foresee with this project is narrowing down the data and my ideas into something meaningful and interesting to visualize. 

## Notes

**Meadow site coordinates provided in metadata** could allow for a site map if interested


**Interesting note from metadata:** "The meadows may have been maintained by Native American burning over at least the past 6,000 years (Highland, 2012). Decimation of Native Americans by the late 1700s and fire suppression since arrival of Europeans in the 1800s is associated with significant contraction of meadow habitat throughout the Cascade Range of Oregon (Miller and Halpern, 1998). In the Andrews Forest, montane meadows have contracted by 45% from 1948 to 2005 (Takaoka and Swanson, 2008, Rice, 2009, Highland, 2012)."